Using Scipion for Stream Image Processing at Cryo-EM Facilities

### 2018-09-26 Review/answer_referee.tex

\end{reviewer}

\begin{reviewer}
\section*{Reviewer 2}

\point{The authors claim that the strength of their software is in its ability to process data while being collected; a point that is not well-addresses by other software. At the same time, they acknowledge that UCSFImage4, Focus, and Relion are software packages similar to Scipion capable of processing data in a high-throughput manner while new data is being acquired at the microscope. A clear differentiation between Scipion and these software packages was lacking in this manuscript. What advantages Scipion provide over the above mentioned software packages?}

\reply %Since there is already a published manuscript describing \scipion, in the present manuscript  we decided to focus on \scipion features that have been specifically developed for working at EM facilities (e.g. streaming). The main reasons why \scipion is better suited than other packages for EM work is rooted in \scipion main design, complimented with the new facilities related features. Nevertheless, 
The reviewer is right in pointing out than a ``A clear differentiation between \scipion and these software packages was lacking in this manuscript''. Therefore, we have decided to add the following subsection to the manuscript

@@

\begin{quote}

\textbf{\scipion Design and Architecture}

We would like to finish this section commenting on some of \scipion  design decisions that differentiate \scipion from other image processing frameworks. \scipion
is fully oriented to integrate existing
EM software and therefore the versatility and number of possible steps
combination is huge. Additionally, \scipion takes care of keeping a 
rigorous traceability and reproducibility as well as to provide an API (application programming interface) to allow the integration  with already existing LIMS systems. At the software architectural level,  the greatest difference 
between \scipion and the other EM image processing packages mentioned in the introduction 
is that  \scipion pays special attention to define abstraction layers to simplify maintenance
and extensibility and follows a strict workflow approach. A few examples will illustrate the main advantages of this approach.

To create a new protocol in \scipion, a single \python file needs to be created. From this file, the system will discover the new protocol and will automatically build
a form and store the protocol related information in a database.
\scipion protocol developers must certainly know \scipion conventions
but they do not need to modify or even understand the existing code.
In  \scipion next version, the abstraction will be increased
since protocols will be implemented as plug-ins,
that is, independent small \python files  that add functionality to the main
\scipion application. The plug-in approach allows parallel development. That is,  
since protocols are implemented independently from the main application
they can be developed and released in parallel by different teams. In this way, support for a new version of \textit{Relion} does not require a new \scipion release but
a release of the independent \textit{Relion} plug-in

\scipion follows a pure workflow approach and therefore, implements pipelines and describes them as a set of interconnected steps (or protocols). In a workflow approach, users (including microscope staff) can reasonably be expected to modify the paths and incorporate new steps using a GUI (graphic user interface). On the contrary, software packages that do not make this separation produce workflows that, in general, are more challenging to create or modify for non programmers.

Another advantage of a clear separation between steps and relationships 
is the improvement of load balance, reproducibility and  provenance (process of tracing and recording the origins of data and its movement between protocols). The first step in a \scipion workflow execution
is to store in a database each of the protocols to be executed as well as how 
outputs of one protocol are attached to the inputs of another protocol. Then, a workflow engine analyzes the stored data and, using as constraint the number of CPUs assigned by the user to the workflow, optimizes the execution, that is, tries to execute as many protocols in parallel as possible. Obviously, the data stored in the database  have enough information to execute each step and therefore guarantee reproducibility and provenance. Additionally, \scipion currently exports a JSON (JavaScript Object Notation) file describing the complete preprocessing workflow used, so that it can be fully reproduced and visualized elsewhere.

@@

\begin{quote}

In the context of free and open source projects it is difficult to create usage statistics that are accurate. But certainly, if we want \scipion to succeed, we need to 
know how frequent the different protocols are used. This is the reason why we have recently developed our \textit{Scipion Usage Data Collector} that monitors -if activated by the user- the usage of the different protocols and send -via HTTP protocol- the information to the developer's team. This information is anonymous and cannot be used to identify the original computer in which \scipion has been executed.

Using the HTTP header it is possible to guess the country in which \scipion has been installed. With these information we have created the usage map and table available at URL \url{http://scipion.i2pc.es/report_protocols/scipionUsage/} and  \url{http://scipion.i2pc.es/report_protocols/protocolTable/}. The data corresponds on ``global'' \scipion usage rather than ``facility'' usage but there 
is a one to one relationship between the  countries with highest number of \scipion projects and the countries with \scipion based EM facilities. The only exception to the rule is Canada where two of the original \scipion developers are setting up a new facility (McGill University).

Note that, in a effort to report real use and skip test projects, the number of projects in the map refers to the number of non empty \scipion projects updated more than once. That is, 
\scipion projects that have been opened in at least two different days. Another constraint is that developers' computers are filtered out.
\end{quote}

### 2018-09-26 Review/1.introduction.tex
to class

Many software packages have been developed to support automated data collection, such as SerialEM \citep{Mastronarde2005}, Leginon \citep{Suloway2009}, UCSF-Tomography \citep{Zheng2007}, Tom Toolbox \citep{Nickell2005}, EPU \citep{EPU}, Latitude in Digital Micrograph \citep{Latitude}, EMMenu \citep{emmenu}, etc. However, the next step,  which is to provide image processing while  data are being collected, is not so well addressed. In fact, in most places this is accomplished through home made scripts \citep[e.g.][]{Pichkur2018}. %(i.e. \citet{Pichkur2018}).
Exceptions are UCSFImage4 \citep{Li2015}, Focus \citep{Biyani2017} and Relion \citep{fernandez2017:relion-pipeline}. As we will show in this article, \scipion differs from these solutions in a larger flexibility, both from the point of view of users and facility staff.

>>>>

Many software packages have been developed to support automated data collection, such as SerialEM \citep{Mastronarde2005}, Leginon \citep{Suloway2009}, UCSF-Tomography \citep{Zheng2007}, Tom Toolbox \citep{Nickell2005}, EPU \citep{EPU}, Latitude in Digital Micrograph \citep{Latitude}, EMMenu \citep{emmenu}, etc. However, the next step,  which is to provide image processing while  data are being collected, is not so well addressed. In fact, in most places this is accomplished through home made scripts \citep[e.g.][]{Pichkur2018}. %(i.e. \citet{Pichkur2018}).
Exceptions are UCSFImage4 \citep{Li2015}, Focus \citep{Biyani2017} and Relion \citep{fernandez2017:relion-pipeline}. As we will show in this article, \scipion differs from these solutions in its larger flexibility, both from the point of view of users and facility staff.

### 2018-09-26 Review/2.overall.tex
to class

We would like to finish this section commenting on some of \scipion architecture design decisions that differentiate \scipion from another image processing frameworks. At the software architectural level,  the greatest difference 
between \scipion and the other EM image processing packages mentioned in the introduction 
is that  \scipion pays special attention to define abstraction layers to simplify maintenance
and extensibility, and that follows a strict workflow approach. A few examples will illustrate the main advantages of this design.

To create a new protocol in Scipion, only a Python script needs to be developed. From this script, the system will discover the new protocol and will automatically build
a form and store the protocol related information in a database.
\scipion protocol developers must certainly know \scipion conventions
but they do not need to modify or even understand the existing code.
In  \scipion next version, the abstraction will be increased
since protocols will be implemented as plug-ins,
that is, independent small python scripts  that add functionality to the main
\scipion application. The plug-in approach allows parallel development. That is,  
since protocols are implemented independently from the main application
they can be developed and released in parallel by different teams. In this way, support for a new version of \textit{Relion} does not require a new \scipion release but
a release of the independent \textit{Relion} plug-in

\scipion follows a pure workflow approach and therefore, implements pipelines and describes them as a set of interconnected steps (or protocols). In a workflow approach, users (including microscope staff) can reasonably be expected to modify the paths and incorporate new steps using a GUI. On the contrary, software packages that do not make this separation produce workflows that, in general, are more challenging to create or modify for non programmers.

Another advantage of a clear separation between steps and relationships 
is the improvement of load valance, reproducibility and  provenance (process of tracing and recording the origins of data and its movement between protocols). The first step in a \scipion workflow execution
is to store in a database each of the protocols to be executed as well as how 
outputs of one protocol are attached to the inputs of another protocol. Then, a workflow engine analyzes the stored data and using as constrain the number of CPUs assigned by the user to the workflow, optimizes the execution, that is, try to execute as many protocols in parallel as possible. Obviously, the data stored in the database  have enough information to execute each step and therefore guaranties reproducibility and provenance.

%for any object created by a  protocol, \scipion always annotate how this object has been generated by detailing: the steps, input data and parameters used in the computation. Data provenance guaranties reproducibility and traceability. %Since provenance is solved by Scipion and not Scipionbox we will not discuss here how it is been achieved. We just note that, before choosing which data has to be stored, it is necessary to define how these data have to be structured so that they can be later recovered and understood . %This abstract modeling of the data also simplifies the software interoperability offered by Scipion.

@@

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently, \scipion offers  five  methods for movie alignment: motioncorr \citep{Li2015}, motioncor2 \citep{Zheng2016:motioncor2}, correlation (an unpublished CPU version that implements in Xmipp a global alignment method and solves it similarly to motioncorr but using robust statistics), unblur \citep{Grant2015:unblur} and optical flow \citep{Abrishami2015:optical-flow}. Then the CTFs of the resulting micrographs are calculated. Scipion provides three CTF estimation methods:
CTFFIND4 (Rohou and Grigorieff, 2015), gCTF (Zhang, 2016) and Xmipp CTF estimation (Sorzano et al., 2007; Vargas et al., 2013). At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that reports up to which frequency two estimations of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. %Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

>>>>

We would like to finish this section commenting on some of \scipion  design decisions that differentiate \scipion from other image processing frameworks. \scipion
is fully oriented to integrate existing
EM software and therefore the versatility and number of possible steps
combination is huge. Additionally, \scipion takes care of keeping a 
rigorous traceability and reproducibility as well as to provide an API (application programming interface) to allow the integration  with already existing LIMS systems. At the software architectural level,  the greatest difference 
between \scipion and the other EM image processing packages mentioned in the introduction 
is that  \scipion pays special attention to define abstraction layers to simplify maintenance
and extensibility and follows a strict workflow approach. A few examples will illustrate the main advantages of this approach.

To create a new protocol in \scipion, a single \python file needs to be created. From this file, the system will discover the new protocol and will automatically build
a form and store the protocol related information in a database.
\scipion protocol developers must certainly know \scipion conventions
but they do not need to modify or even understand the existing code.
In  \scipion next version, the abstraction will be increased
since protocols will be implemented as plug-ins,
that is, independent small \python files  that add functionality to the main
\scipion application. The plug-in approach allows parallel development. That is,  
since protocols are implemented independently from the main application
they can be developed and released in parallel by different teams. In this way, support for a new version of \textit{Relion} does not require a new \scipion release but
a release of the independent \textit{Relion} plug-in

\scipion follows a pure workflow approach and therefore, implements pipelines and describes them as a set of interconnected steps (or protocols). In a workflow approach, users (including microscope staff) can reasonably be expected to modify the paths and incorporate new steps using a GUI (graphic user interface). On the contrary, software packages that do not make this separation produce workflows that, in general, are more challenging to create or modify for non programmers.

Another advantage of a clear separation between steps and relationships 
is the improvement of load balance, reproducibility and  provenance (process of tracing and recording the origins of data and its movement between protocols). The first step in a \scipion workflow execution
is to store in a database each of the protocols to be executed as well as how 
outputs of one protocol are attached to the inputs of another protocol. Then, a workflow engine analyzes the stored data and, using as constraint the number of CPUs assigned by the user to the workflow, optimizes the execution, that is, tries to execute as many protocols in parallel as possible. Obviously, the data stored in the database  have enough information to execute each step and therefore guarantee reproducibility and provenance. Additionally, \scipion currently exports a JSON (JavaScript Object Notation) file describing the complete preprocessing workflow used, so that it can be fully reproduced and visualized elsewhere.

@@

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently, \scipion offers  five  methods for movie alignment: motioncorr \citep{Li2015}, motioncor2 \citep{Zheng2016:motioncor2}, correlation (an unpublished CPU version that implements in Xmipp a global alignment method and solves it similarly to motioncorr but using more robust statistics), unblur \citep{Grant2015:unblur} and optical flow \citep{Abrishami2015:optical-flow}. Then the CTFs of the resulting micrographs are calculated. \scipion provides three CTF estimation methods:
CTFFIND4 \citep{Rohou2015}, gCTF \citep{Zhang2016} and Xmipp CTF estimation \citep{sorzano2007, Vargas2013}. At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that reports up to which frequency two estimations of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. %Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

### 2018-09-26 Review/3.workflow.tex
to class

\scipion original design satisfied CNB requirements, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm), where a rigorous booking system is followed. There, middleware (software that acts as a bridge between applications) was developed to interface with this booking platform to produce better reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond,  where projects are handled by ISPyB (a customized laboratory information management system -LIMS- \citep{Delageniere2011:ispb}) and data are saved on a distributed file system. % that heavily penalizes access to lists of files. 
In this environment, \scipion needs to constantly interchange information with ISPyB and with other applications through a message queue system called \emph{ActiveQ}.% and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created.

>>>>

\scipion original design satisfied CNB requirements, in which \scipion automatically fetches newly recorded movies from a network mounted disk. After this first installation, \scipion was deployed on SciLifeLab (a centralized facility located at Stockholm), where a rigorous booking system is followed. There, middleware (software that acts as a bridge between applications) was developed to interface with this booking platform to produce more precise reports. Lately,  \scipion has been installed in  large synchrotrons  such as ESRF and Diamond,  where projects are handled by ISPyB (a customized laboratory information management system -LIMS- \citep{Delageniere2011:ispb}) and data are saved on a distributed file system. % that heavily penalizes access to lists of files. 
In this environment, \scipion needs to constantly interchange information with ISPyB and with other applications through a message queue system called \emph{ActiveQ}.% and it cannot longer check if there is a new file in a particular storage device but must wait to be informed by the system that a new file has been created.

### 2018-09-26 Review/3.workflow_diamond.tex
to class

\subsection{Case 3: Scipion at eBIC (Diamond Light Source)}

@@

\subsection{Additional Software Developments: ISPyB monitor, Scipion Headless}

@@

Unlike CNB and SciLifeLab cryo-EM services, where the
Scipion default monitor system displays the progress of the automatic image
processing, in Diamond, the electronic notebook ISPyB is the natural framework
to record and show metadata about the different experiments. A new monitor has thus been developed to allow Scipion to directly interface with ISPyB using the ISPyB python API.
%To monitor the progress of the automatic image processing, CNB and \scilifelab cryo-EM services make use of \scipion default monitor system. However, in Diamond, the electronic notebook ISPyB is the standard tool to record and show metadata about the different experiments. 
%In this way, a new monitor has been developed which, through a ISPyB python API, allows \scipion to directly interface with ISPyB.  
This setup provides the users with a web interface showing general metadata about their experiment, as well as  real-time updates on steps of the process and relevant output, such as motion corrected images, drift values, CTF images, etc.

>>>>

\subsection{Case 3: \scipion at eBIC (Diamond Light Source)}

@@

\subsection{Additional Software Developments: ISPyB monitor, \scipion Headless}

@@

Unlike CNB and SciLifeLab cryo-EM services, where the
Scipion default monitor system displays the progress of the automatic image
processing, in Diamond, the electronic notebook ISPyB is the natural framework
to record and show metadata about the different experiments. A new monitor has thus been developed to allow \scipion to directly interface with ISPyB using the ISPyB python API.
%To monitor the progress of the automatic image processing, CNB and \scilifelab cryo-EM services make use of \scipion default monitor system. However, in Diamond, the electronic notebook ISPyB is the standard tool to record and show metadata about the different experiments. 
%In this way, a new monitor has been developed which, through a ISPyB python API, allows \scipion to directly interface with ISPyB.  
This setup provides the users with a web interface showing general metadata about their experiment, as well as  real-time updates on steps of the process and relevant output, such as motion corrected images, drift values, CTF images, etc.

### 2018-09-26 Review/3.workflow_scilifelab.tex
to home
to class

\subsection{Case 2: Scipion at the Swedish National Cryo-EM Facility}

@@

At the beginning of 2016, the Cryo-EM facility at \scilifelab became one of the early adopters of \scipion streaming processing. Although the microscope operators were using home-made scripts to perform motion correction (with motioncor2) and CTF estimation (with Gctf), they recognized the importance of a more general
framework. In particular, they were interested in the possibility of modifying easily the processing workflow as well as provide users and staff with  graphical tools for data analysis and quick feedback regarding the data collection.
Although initially the same setup script used at
CNB was adopted, a new one was developed to fully satisfy the facility
requirements. In the following sections we will briefly describe the computational infrastructure at the \scilifelab node and the implementation of a Session Wizard to automate the initial setup while fetching information from other sources external to \scipion. 
%[** thalos falcon III, titan k2 and falcon III]

@@

\subsubsection{Additional Software Developments: \emph{Session Wizard}}
In order to interface \scipion with the microscope infrastructure an upgraded version of the original script developed at CNB was used. Unfortunately, it was difficult to reuse the existing code due to a different folder structure and, more important, a different approach to user and microscope management. To address the new set of requirements, a new wizard has been development at \scilifelab. (In this contest wizard refers to a helper script which guide users through a certain process.) %For this new wizard, a basic data model was designed representing the current processing and data organization at \scilifelab. %The model is composed by the following entities:

>>>>

\subsection{Case 2: \scipion at the Swedish National Cryo-EM Facility}

@@

At the beginning of 2016, the Cryo-EM facility at \scilifelab became one of the early adopters of \scipion streaming processing. Although the microscope operators were using home-made scripts to perform motion correction (with motioncor2) and CTF estimation (with Gctf), they recognized the importance of a more general
framework. In particular, they were interested in the possibility of modifying easily the processing workflow as well as provide users and staff with  graphical tools for data analysis and quick feedback regarding the data collection.
Although initially the same setup script used at
CNB was adopted, a new one was developed to fully satisfy the facility
requirements. In the following sections we will briefly describe the computational infrastructure at the \scilifelab node and the implementation of a Session Wizard to automate the initial setup while fetching information from other sources external to \scipion. %(In this contest wizard refers to a helper script which guide users through a certain process.)
%[** thalos falcon III, titan k2 and falcon III]

@@

\subsubsection{Additional Software Developments: \emph{Session Wizard}}
In order to interface \scipion with the microscope infrastructure, an upgraded version of the original script developed at CNB was used. Unfortunately, it was difficult to reuse the existing code due to a different folder structure and, more important, a different approach to user and microscope management. To address the new set of requirements, a new wizard has been development at \scilifelab. %(In this contest wizard refers to a helper script which guide users through a certain process.) %For this new wizard, a basic data model was designed representing the current processing and data organization at \scilifelab. %The model is composed by the following entities:

### 2018-09-26 Review/4.future.tex
to class

As part of the EOSCPilot project (\url{https://eoscpilot.eu/cryoem}), we are also improving  how \scipion exports workflows in order to ease reporting of the work done. The goal is to write a workflow file that fully describes the image processing steps, so that if the same input data is provided to the workflow the same results should be obtained. In this way we want to make  the data produced by \scipion more compliant with FAIR guidelines given by the \citet{eu2016:fair}. This workflow file could go with the raw data acquired by CryoEM facilities and deposited at common EM databases such as EMPIAR or EMDB \citep{Patwardhan2016:databasesEM}. To facilitate the visualization of the workflow file in any of these databases, we are  developing a  webcomponent (\scipion \emph{workflow viewer}) that will easily allow these repositories to visualize the workflow on their web pages (\url{https://github.com/I2PC/web-workflow-viewer}). Finally, we are implementing a workflow repository. This repository will contain ready-to-use workflows that support a range of use cases.

>>>>

As part of the EOSCPilot project (\url{https://eoscpilot.eu/cryoem}), we have improved 
how \scipion exports workflows in order to ease reporting of the work done. In this way we want to make  the data produced by \scipion more compliant with the FAIR guidelines given by the \citet{eu2016:fair}. This workflow file could go with the raw data acquired by CryoEM facilities and deposited at common EM databases such as EMPIAR or EMDB \citep{Patwardhan2016:databasesEM}. 
%As part of the EOSCPilot project (\url{https://eoscpilot.eu/cryoem}), we are also improving  how \scipion exports workflows in order to ease reporting of the work done. The goal is to write a workflow file that fully describes the image processing steps, so that if the same input data is provided to the workflow the same results should be obtained. In this way we want to make  the data produced by \scipion more compliant with FAIR guidelines given by the \citet{eu2016:fair}. 
To facilitate the visualization of the workflow file in any of these databases, we are  developing a  webcomponent (\scipion \emph{workflow viewer}) that will easily allow these repositories to visualize the workflow on their web pages (\url{https://github.com/I2PC/web-workflow-viewer}). Finally, we are implementing a workflow repository. This repository will contain ready-to-use workflows that support a range of use cases. A preliminary neta version of the workflows repository is available at \url{http://workflows.scipion.i2pc.es/}

### 2018-09-26 Review/6.acknowledgments.tex
to class

The Spanish Ministry of Economy and Competitiveness through Grants BIO2016-76400-R(AEI/FEDER, UE) and AEI/FEDER BFU 2016 74868P, the Comunidad Aut\'{o}noma de Madrid through Grant: S2017/BMD-3817, European Union (EU) and Horizon 2020 through grant iNEXT (INFRADEV-1-2014-1, Proposal: 654248). The ``Knut \& Alice Wallenberg Foundation'', and ``A Pilot Facility development grant from Science for Life Laboratory''. European Union (EU) and Horizon 2020 through grant EOSCpilot (INFRADEV-04-2016, Proposal: 739563).

>>>>

The Spanish Ministry of Economy and Competitiveness through Grants BIO2016-76400-R(AEI/FEDER, UE) and AEI/FEDER BFU 2016 74868P, the Comunidad Aut\'{o}noma de Madrid through Grant: S2017/BMD-3817, European Union (EU) and Horizon 2020 through grant Corbel (INFRADEV-1-2014-1, Proposal: 654248). The ``Knut \& Alice Wallenberg Foundation'', and ``A Pilot Facility development grant from Science for Life Laboratory''. European Union (EU) and Horizon 2020 through grant EOSCpilot (INFRADEV-04-2016, Proposal: 739563).
